---
title: "PS270 Final Project"
author: Shuzhi Liu
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

The data source I used to complete my Final Project is the American National Election Study (ANES) – a long-running project that conducts in-depth surveys on voting behavior, public opinion, and political participation in the United States. My research question is: Is economic insecurity an indicator of the likelihood of being identified as conservative ideologically in the United States? Based upon economic anxiety and threat response theory I hypothesize that those who report experiencing, or believe they are experiencing economic insecurity, will have a greater tendency to identify themselves as conservative ideologically. A possible behavioral cause for this may be that economic uncertainty leads to a preference for social order, stability and traditional institutions; all characteristics typically associated with conservative ideologies. My key independent variable is economic insecurity which was measured through survey questions such as respondents' reported level of satisfaction regarding their finances or whether or not respondents believed that the nation's economic situation is improving or deteriorating. This allows for variation across each respondent thereby permitting regression analysis. My dependent variable is respondents' ideological self-identification which is measured by respondents' placement on the GSS liberal-conservative scale (which generally ranges from extremely liberal to extremely conservative). If support for my hypothesis is found in a regression model where high levels of economic insecurity are related to high (i.e. more conservative) placements on the ideology scale, then this will confirm my hypothesis. Conversely, if there is a significant relationship between economic insecurity and liberal ideological identity or if there is no statistically significant relationship between economic insecurity and ideological identity — either liberal or conservative — then this will disconfirm my hypothesis.

```{r}
library(tidyverse)
```


```{r}
anes <- read.csv("data/anes_timeseries_2024_csv_20250808.csv")

proj_data <- anes |>
  select(V241177, V241539) |>
  mutate(
    ideology = as.numeric(V241177),
    fin_worry = as.numeric(V241539),
    fin_worry_rev = 6 - fin_worry
  ) |>
  filter(
    ideology >= 1 & ideology <= 7,
    fin_worry >= 1 & fin_worry <= 5
  )

plot_data <- proj_data |>
  group_by(fin_worry_rev) |>
  summarize(
    mean_ideology = mean(ideology),
    n = n()
  )

line_graph <- ggplot(plot_data, aes(x = fin_worry_rev, y = mean_ideology)) +
  geom_point(size = 3) +
  geom_line() +
  scale_x_continuous(
    breaks = 1:5,
    labels = c(
      "Not at all worried",
      "A little worried",
      "Moderately worried",
      "Very worried",
      "Extremely worried"
    )
  ) +
  labs(
    title = "Average Ideology by Level of Financial Worry",
    x = "Current financial worry",
    y = "Average ideology (higher = more conservative)"
  )

line_graph
```

